Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Fixes in formulae

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems 
with multiple data sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
reveal source-specific clusters instead of grouping data instances of the same class, regardless of data sources. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em data set shift}~\cite{data set_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

@@

\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq \i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:gaussian_kernel}
\end{equation}

@@

\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} d(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

@@

Additional experiments with embedding secondary data sets from independent
studies are provided in the accompanying Python notebooks. We embedded brain
cells from \cite{} to the reference embedding shown in
Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 
of X\% and Y\%. \todo{Add errors}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbours search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; M reference, N secondary cells, X min N-search,
Y min optimization). \todo{Add nunmbers}

We demonstrate a case where nearest neighbor-based positioning is insufficient
in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
optimal positioning using the t-SNE loss function is much more dispersed and
rightfully shows a more considerable variation in the secondary data. 
Some points may also fall into the empty regions, while after optimization they typically
move closer to same-typed groups. While the nearest neighbour-based approach
always places new data points to their most similar cluster, our proposed
t-SNE approach can identify extreme outliers and place them far from the
reference cluster.

>>>>

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems 
with multiple data sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
reveal source-specific clusters instead of grouping data instances of the same class, regardless of data sources. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

@@

\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:gaussian_kernel}
\end{equation}

@@

\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(\max \{ N, M \})$.

@@

Additional experiments with embedding secondary data sets from independent
studies are provided in the accompanying Python notebooks. We embedded brain
cells from~\cite{} to the reference embedding shown in
Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 
of X\% and Y\%. \todo{Add errors}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbors search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; M reference, N secondary cells, X min N-search,
Y min optimization). \todo{Add nunmbers}

We demonstrate a case where nearest neighbor-based positioning is insufficient
in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
optimal positioning using the t-SNE loss function is much more dispersed and
rightfully shows a more considerable variation in the secondary data. 
Some points may also fall into the empty regions, while after optimization they typically
move closer to same-typed groups. While the nearest neighbor-based approach
always places new data points to their most similar cluster, our proposed
t-SNE approach can identify extreme outliers and place them far from the
reference cluster.
